---
title: "HELPmiss (Data Verbs) Activity"
author: "Neeraj Thokala"
output: html_notebook
---


## Set Up:

```{r message=FALSE}
rm(list = ls())     # clean up your R environment

# load packages
library(tidyverse)  # includes lots of data verbs like `group_by()` and `summarise()`
library(mosaicData) # includes the `HELPmiss` data set 


# Load the `HELPmiss` data set into our RStudio environment
data("HELPmiss", package = "mosaicData")

```

## Helpful links: 

- Look though the DC Textbook for `tidyverse` functions. 
- Check out some of these RStudio cheat sheets:
  - <https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf> 
  - <https://www.rstudio.com/resources/cheatsheets/>
  - <https://s3.amazonaws.com/assets.datacamp.com/blog_assets/Tidyverse+Cheat+Sheet.pdf>
  



## Task 1: Data Description

The HELP study was a clinical trial for adult patients recruited from a detoxification unit. Patients with no primary care physician were randomized to receive a multidisciplinary assessment and a brief motivational intervention or usual care, with the goal of linking them to primary medical care.

Subjects were interviewed at baseline during their detoxification stay and follow-up interviews were undertaken every 6 months for 2 years. A variety of continuous, count, discrete, and survival time predictors and outcomes were collected at each of these five occasions.

This dataset is a superset of the HELPrct data with 17 subjects with partially observed data on some of the baseline variables. This is a subset of the HELPfull data which includes 5 timepoints and many additional variables.



## Task 2: Basic Summaries 

*Produce one or more R expressions involving `summarize()` and `HELPmiss` to address each of the following prompts.*

1. number of people (cases) in `HELPmiss` study

```{r}
HELPrct %>%
  summarise( total_people = n()) 

```

2. total number of times in the past 6 months entered a detox program (measured at baseline) for all the people in `HELPmiss`. 

```{r}
HELPrct %>%
  summarise( total_people = n(),total_detox= sum(e2b,na.rm=TRUE)) 

```


3. mean time (in days) to first use of any substance post-detox for all the people in `HELPmiss`


```{r}
HELPrct %>%
  summarise( use_mean = mean(daysanysub,na.rm=TRUE) )

```



## Task 3: Group Summaries  

*Repeat task 2 above, but add code chunks to calculate result group-by-group according to each prompt (i.e. each promt should have three statements for each of the three prompts in task 2).  Be sure to show all R code and write a sentence or two about what you observe in the results. Remember, you can add multiple statistics inside `summary`.* 


- males versus females
- homeless or not
- substance
- break down the homeless versus housed further, by sex
- homeless versus housed broken down by substance

### males versus females
```{r}
HELPrct %>%
  group_by(female)%>%
  summarise( total_people = n(),total_detox= sum(e2b,na.rm=TRUE),use_mean = mean(daysanysub,na.rm=TRUE) ) 

```

### homeless or not
```{r}
HELPrct %>%
  group_by(homeless)%>%
  summarise( total_people = n(),total_detox= sum(e2b,na.rm=TRUE),use_mean = mean(daysanysub,na.rm=TRUE) )

```


### substance
```{r}
HELPrct %>%
  group_by(substance)%>%
  summarise( total_people = n(),total_detox= sum(e2b,na.rm=TRUE),use_mean = mean(daysanysub,na.rm=TRUE) ) 

```

### homeless versus housed broken down by sex
```{r}
HELPrct %>%
  group_by(female,homeless)%>%
  summarise( total_people = n(),total_detox= sum(e2b,na.rm=TRUE),use_mean = mean(daysanysub,na.rm=TRUE) ) 
```

### homeless versus housed broken down by substance
```{r}
HELPrct %>%
  group_by(homeless,substance)%>%
  summarise( total_people = n(),total_detox= sum(e2b,na.rm=TRUE),use_mean = mean(daysanysub,na.rm=TRUE) ) 

```



## Task 4: Data Visualization & Observations

*Include one or more interesting plots from this data set involving at least 3 variables per plot. Write a few sentences to explain the story that your plot tells about these data.  You can expand on the relationships that you studied in Task 2, or you can explore a different group of variables in `HELPmiss` that show something interesting. Remember to use the interactive commands in the console, generate the R commands that will reproduce your plot, and then paste the R commands into an R chunk in the RMarkdown file.*  

*Remember, you can use* `esquisser` *or* `mplot` *in your console. But only include the ggplot code in this Rmd document.*


```{r}
gf_point(daysanysub~mcs, color=~racegrp,data=HELPrct)

```



